Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars

### 2014-10-20 drafts/cfp/draft.tex
another pass through the draft

\section{Abstract}
How can machines learn about a system of concepts from sentences about
those concepts? One approach is to estimate the distribution from
which those sentences are drawn, and, in doing so, infer the latent
concepts and relationships that best explain the data. We present
early work on a learning framework called Latent Predicate Networks
(LPNs) which attempts to learn concepts by inferring the parameters of
probabilistic context-sensitive grammars that best describe sentences.
We show that for a small fragment of sentences expressing
relationships between English number words, we can use hierarchical
Bayesian inference to learn LPNs that can answer simple queries about
previously unseen linguistic concepts within this domain. These
generalizations, we argue, demonstrate LPNs' promise as a tool for
learning and representing conceptual knowledge in language.

% A popular approach in statistical machine learning is to automatically
% find the latent \emph{features} of the data that are most relevant or
% informative. But a latent feature based view of concept learning does
% not match up naturally with the structured symbolic knowledge that
% seems to underlie people's understanding of many domains. How can we
% combine the flexibility, noise tolerance, and computational
% practicality of latent feature based statistical machine learning with
% the symbolic knowledge representations that best capture human
% understanding? Here present a learning model we dub Latent Concept
% Grammars (LCGs) as a proposal. An LCG is a densely connected set of
% probabilistic string rewrite rules over an initially uninterpreted set
% of predicates. When fit to data using a hierarchical Bayesian prior,
% these predicates assume the roles of informative latent concepts. This
% approach allows us to learn knowledge that is stochastic and
% recursively structured while maintaining computational
% tractability. We show how learning in LCGs can be implemented using
% the probabilistic logic programming system PRISM, and we present
% preliminary experiments.
}
% How do people learn systems of related concepts from
% data?  Consider, for example, a school-age child's knowledge of
% number. This knowledge supports a large number of related operations:
% the child knows which words denote numbers, how to use those words to
% label cardinalities of sets, how to translate between digit
% representations and word representations of number, how to compare
% cardinalities, etc. The child knows how to perform these operations
% over an unbounded set of elements, even though the data that underlies
% this knowledge is small and noisy. Here we propose that such systems
% of related concepts can be learned via hierarchical probabilistic
% inference over probabilistic string rewrite systems (PSRS) -- a mildly
% context-sensitive class of probabilistic string gramamrs that
% corresponds to a restricted fragment of stochastic logic programs
% (SLPs). Specifically, we extend the variational
% expectation-maximization algorithm for Hierarchical Dirichlet PCFGs to
% accomodate PSRSs. This algorithm we present enables a tractible
% compromise between highly restrictive context-free knowledge
% representations and the intractability of performing similar inference
% procedures over general logic programs. Empirically, we show how such
% restricted string rewrite systems can capture non-trivial conceptual
% knowledge and how our algorithm is able to learn such knowledge from
% data.

% \section{Abstract 2}
% What does it mean to understand a concept, and how do people come to
% such an understanding.

@@

As Claude Shannon demonstrated with his guessing game experiment in
1950~\cite{Shannon1950}, an adult English speaker's understanding of
her language and her world allows her to reveal that the entropy of a
character in English text is only a little more than one
bit. Plausibly, any machine learning system that were able to predict
natural language text with such accuracy will have learned much about
the concepts expressed in that text. This intuition motivates us to
invert Shannon's experiment and attempt to learn about linguistic
concepts by learning a probabilistic generative model for natural
language.

Probabilistic Context Free Grammars (PCFGs) are popular and much
studied generative models for natural language
syntax~\cite{Someone}. And though there ongoing debate about their
ability to express the syntax of natural language, there is no doubt
that they cannot express dependencies between words that depend upon
their meaning. Consider an example from the domain of sentences
addressed in this paper: ``after twenty eight comes twenty nine.''
The second ``twenty'' depends on the first, and the ``nine'' depends
on the ``eight,'' and this dependency pattern is a clear example of a
\emph{cross-serial dependency}, a dependency pattern that context-free
grammars do not support.

Our approach, then, will be to explore the use of probabilistic
context \emph{sensitive} grammars for modeling natural language,
extending the tools of hierarchical grammar induction~\ref{Cite} from
context-free grammars to ones that can express not only syntactic
relationships but semantic ones as well. In the place of latent
syntactic categories, we will refer to latent predicates -- relations
between sentence components -- that, when learned, should correspond
to the concepts that underlie sentence semantics.

@@

An LPN is a hierarchical Bayesian model of strings. It is an
extension of the Hierarhical Dirichlet PCFG model (HD-PCFG) to
Probabilistic Range Concatenation Grammars (PRCGs).

@@

We extend RCGs to Probabilistic Range Concatenation Grammars (PRCGs)
by annotating each clause $C_k \in P$ with probabilities $p_k$ such
that for all predicates ${A \in N, \, \sum_{k:head(C_k)=A} p_k = 1}$. A
PRCGs defines a distribution over strings $x$ by sampling from
derivations of $S(x)$ according to the product of probabilities of
clauses used in that derivation. This is a well defined distribution
as long as no probability mass is placed on derivations of infinite
length; in this paper, we only consider PRCGs with derivations of
finite length, so we need not worry about this requirement.

@@

An LPN is an PRCG with $K$ binary predicates $A_1, \dots, A_K$ with a
densely connected structure. The architecture of the LPNs used in this
work is described in Figure~\ref{fig:architecture}. Specifically, for
each latent predicate $A_k$,

\begin{figure}[t]
		\includegraphics[width=\linewidth]{lpn/lpn.pdf}
		\caption{A schematic LPN where $m = |L|,\, n = |N|-1$,\, arity$(A_i) = 2$, and arity$(S) = 1$.}
		\label{fig:architecture}
\end{figure}

@@

\subsection{Related Work}
Johnson and Liang, we are interested in inferring the most likely
grammar given some collection of sentences, and in doing so without a
pre-specified limit on the complexity of the grammar. Like Cussens, we
are interested in efficiently extending this algorithm to more
powerful and expressive formalisms than CFGs.

@@

\section{Inference}
Inference in hierarchical Dirichlet processes in stochastic grammars
and stochastic logic programs has been an active area of research in
recent
decades~\cite{DBLP:conf/emnlp/LiangPJK07,goldwater2006contextual,johnson2006adaptor,cussens2001parameter}.
Variational inference is a popular approach in this domain and it is
the approach we adopt here. As we explain in the next section, we
implemented inference by translating LPNs into PRISM programs, and we
use the builtin Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}.

@@

\begin{figure}[t]
	\centering
	\begin{minipage}[b]{\linewidth}
A(a, a) :: 0.25. \\
A(b, b) :: 0.3. \\
A(X a, Y a) $\leftarrow$ A(X, Y) :: 0.25. \\
A(X b, Y b) $\leftarrow$ A(X, Y) :: 0.2.
		\subcaption{}
		\label{fig:grammar}
	\end{minipage}
	%\rule[0pt]{0.8\linewidth}{.5pt}
	\begin{minipage}[b]{\linewidth}
		\begin{lstlisting}
values('A', [1, 2, 3, 4],
    [0.25, 0.30, 0.25, 0.20]).

@@

LPNs can be encoded as a restricted and tractable subclass of Stochastic Logic Programs (SLP). Specifically, we encode LPNs as programs in PRISM PROLOG. PRISM provides probabilistic execution and Bayesian inference over SLPs.

To illustrate the encoding, consider the 2-copy language (CL2) in Figure \ref{fig:copy}. CL2 is a toy language producing two identical strings of $a$s and $b$s, making it mildly context-sensitive. Here, we see four rules concerning a single symbol, $A$, whose total probability sums to unity.

To translate LPNs into PRISM SLPs, we first define the distribution over each symbol. We use the {\tt values} predicate, which takes an atom, in this case the name of the symbol, the values that atom can take on, and the probability of each value. For inference, we initialize the distribution uniformly. For our encoding, each possible value for an atom (or symbol) corresponds to exactly one clause in CL2 where that symbol appears on the left-hand side. There are four clauses for $A$ in CL2, so {\tt values} can take on one of four values, each of whose probabilities corresponds to a rule in CL2.

For each rule in the LPN, we also define a corresponding clause of the {\tt reduce} predicate in our PRISM program. These {\tt reduce} rules are where the actual relations between the left-hand and right-hand side of each rule are encoded. Because PROLOG isn't perfectly declarative, the actual {\tt reduce} clauses are slightly more complex than presented here, reversing the order of the {\tt append} and recursive {\tt lpn} calls based on whether we are sampling or parsing.

To parse or sample a sentence, we call {\tt lpn(P-IN)}, where {\tt P} is the top-level symbol, {\tt IN} is a list of terminal lists, and {\tt -} denotes a pair ({\it e.g.} {\tt lpn('A'-[[a,a,
b,a],[a,a,b,a]])}). A sample or parse begins by probabilistically choosing some rule for the left-hand side via {\tt msw(P,V)} and then recursively reducing the input term using the chosen rule ({\tt reduce(P-IN,V)}). 
\section{Experiments}

@@

%1) Introduction
%  - psychological motivation 
%    - juxtaposing the kinds of knowledge that freebase has and the kinds that children learn
%    - compositional semantic structure --> derive semantics from syntax
%  - what is the contribution?
%     - the use of string rewrite formalisms to capture conceptual knowledge and conceptual learning
%     - to bring tools from computational linguistics to learning semantics
%  - what kind of concept systems are we trying to capture?
%  - why restrict the formalism? 
%     - tractability of inference 
%  - why go beyond context free grammars? 
%     - if they are intractable to learn anyway, why try to go beyond them? 
%  - Related Work
%2) Model
%- stochastic linear context free rewrite systems
%  - model: what is the data? and what is being inferred? what is the prior? 
%  - this can be written as PRISM program, or an SLP. 
%  - inference: variational approximation 
%3) Experiments
%  a) worked example
%  b) Number system (main result)
%  3b) Results
%4) Conclusion
%5) Citations

>>>>

{\small For humans, learning abstract concepts and learning language
  go hand in hand: we acquire abstract knowledge primarily through
  linguistic experience, and acquiring abstract concepts is a crucial
  step in learning the meanings of linguistic expressions. How,
  though, can machines learn systems of concepts from sentences about
  those concepts? One approach is to estimate the distribution from
  which those sentences are drawn, and, in doing so, infer the latent
  concepts and relationships that best explain the data. We present
  early work on a learning framework called Latent Predicate Networks
  (LPNs) which learns concepts by inferring the parameters of
  probabilistic context-sensitive grammars over sentences.  We show
  that for a small fragment of sentences expressing relationships
  between English number words, we can use hierarchical Bayesian
  inference to learn LPNs that can answer simple queries about
  previously unseen linguistic concepts within this domain. These
  generalizations demonstrate LPNs' promise as a tool for learning and
  representing conceptual knowledge in language.}

@@

As Claude Shannon demonstrated with his guessing game experiment in
1950~\cite{Shannon1950}, an adult English speaker's understanding of
her language and her world is such that the entropy of a character in
English text is slightly more than one bit. Plausibly, any machine
learning system able to predict natural language text with such
accuracy will have learned much about the concepts expressed in that
text. This intuition motivates us to invert Shannon's experiment and
learn linguistic concepts by learning a probabilistic generative model
for natural language.

Probabilistic Context Free Grammars (PCFGs) are popular and much
studied generative models for natural language
syntax~\cite{Someone}. While debate continues about their ability to
express the syntax of natural language, it is agreed that they cannot
express dependencies between words that depend upon each other for
their meaning. Consider an example from the domain of sentences
addressed in this paper: ``after twenty eight comes twenty nine.''
The second ``twenty'' depends on the first ``twenty'' and the
``eight'', and the ``nine'' depends on the ``eight''. This pattern is
a clear example of the sort of \emph{cross-serial dependencies} which
context-free grammars cannot support.

Our approach, then, will be to explore the use of probabilistic
context \emph{sensitive} grammars for modeling natural language,
extending the tools of hierarchical grammar induction~\cite{SOMEONE} from
context-free grammars to ones that can express not only syntactic
relationships but semantic ones as well. In the place of latent
syntactic categories, we will refer to latent predicates -- relations
between sentence components -- that, when learned, should correspond
to the concepts that underlie sentence semantics.

@@

An LPN is a hierarchical Bayesian model of strings
extending the Hierarchical Dirichlet PCFG model (HD-PCFG) to
Stochastic Range Concatenation Grammars (SRCGs).

@@

We extend RCGs to SRCGs by annotating each clause $C_k \in P$ with
probabilities $p_k$ such that for all predicates ${A \in N, \,
  \sum_{k:head(C_k)=A} p_k = 1}$. An SRCG defines a distribution over
strings $x$ by sampling from derivations of $S(x)$ according to the
product of probabilities of clauses used in that derivation. This is a
well defined distribution as long as no probability mass is placed on
derivations of infinite length; in this paper, we only consider SRCGs
with derivations of finite length, so we need not worry about this
requirement.

@@

An LPN is an SRCG with 1 unary predicate $S$, and $K$ densely
connected binary predicates $A_1, \dots, A_K$ (Figure
\ref{fig:architecture}).  Specifically, for each latent predicate
$A_i$, we define rules such that $A_i(w_a,w_b)$ is true for each
possible pair of terminals, $w_a,w_b \in T$. We also define rules such
that $A_i(XY,UV)$ is true for each possible pair of predicates,
$A_j,A_k$, and every possible ordering of the variables $X,Y,U$ and
$V$ across $A_j$ and $A_k$. Finally, we define $S(XY)$ to be the
concatenation of two strings generated by $A_1$.

\begin{figure}[t]
		\includegraphics[width=\linewidth]{lpn/lpn.pdf}
		\caption{A schematic LPN.}
		\label{fig:architecture}
\end{figure}

@@

@@

\section{Inference}
Inference in hierarchical Dirichlet processes in stochastic grammars
and stochastic logic programs (SLPs) has been an active area of research in
recent
decades~\cite{DBLP:conf/emnlp/LiangPJK07,goldwater2006contextual,johnson2006adaptor,cussens2001parameter}.
Variational inference is a popular approach in this domain and it is
the approach we adopt here. As we explain in the next section, we
implemented inference by translating LPNs into PRISM programs, and we
use the builtin Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}.

@@

\begin{figure}[t]
	\centering
	\begin{minipage}[b]{\linewidth}
1) A(a, a) :: 0.25. \\
2) A(b, b) :: 0.3. \\
3) A(X a, Y a) $\leftarrow$ A(X, Y) :: 0.25. \\
4) A(X b, Y b) $\leftarrow$ A(X, Y) :: 0.2.
		\subcaption{}
		\label{fig:grammar}
	\end{minipage}
	%\rule[0pt]{0.8\linewidth}{.5pt}
	\begin{minipage}[b]{\linewidth}
		\begin{lstlisting}
values('A', [1, 2, 3, 4],
    [0.25, 0.30, 0.25, 0.20]).

@@

LPNs can be encoded as a restricted and tractable subclass of
SLPs. Specifically, we encode LPNs with PRISM PROLOG to obtain
built-in predicates for probabilistic execution and Bayesian inference
over SLPs (Figure \ref{fig:copy}).

First, we define the distribution over each non-terminal symbol, $n
\in N$, using the {\tt values} predicate. For each $n$ we define the
values the symbol can take on ({\it i.e.} one for each rule), as well
as the probability of each value. For each rule in the LPN, we also
define a corresponding clause of the {\tt reduce} predicate to capture
the actual relations between the left-hand and right-hand side of each
rule. Sampling and parsing begins with a call to {\tt lpn(P-IN)},
which chooses some rule for the left-hand side via {\tt msw(P,V)} and
recursively reduces the input term using the chosen rule ({\tt
reduce(P-IN,V)}).

\section{Experiments}

@@
